---
title: "Hands-on Exercise 7"
author: "Meng Xinyu"
---

# ***27  Chapter 27: Modelling, Visualising and Analysing Network Data with R***

```{r}
#27.2 Getting Started
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts, 
               concaveman, ggforce)
```

```{r}
#27.3 The Data
GAStech_edges <- read_csv("data/GAStech-email_edges.csv")
GAStech_nodes <- read_csv("data/GAStech_email_nodes.csv")
```

```{r}
#27.3.4 Reviewing the imported data
glimpse(GAStech_edges)
```

```{r}
#27.3.5 Wrangling time
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

```{r}
#27.3.7 Wrangling attributes
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## ***27.4 Creating network objects using tidygraph***

```{r}
#27.4.3 Using tbl_graph() to build tidygraph data model.
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
#27.4.4 Reviewing the output tidygraph’s graph object
GAStech_graph
```

```{r}
#27.4.6 Changing the active object
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## **27.5 Plotting Static Network Graphs with ggraph package**

```{r}
#27.5.1 Plotting a basic network graph
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

```{r}
#27.5.2 Changing the default network graph theme
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

```{r}
#27.5.3 Changing the coloring of the plot
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

```{r}
#27.5.5 Fruchterman and Reingold layout
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

```{r}
#27.5.6 Modifying network nodes
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

```{r}
#27.5.7 Modifying edges
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

```{r}
#27.6.1 Working with facet_edges()
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

```{r}
#27.6.2 Working with facet_edges()
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

```{r}
#27.6.3 A framed facet graph
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

```{r}
#27.6.4 Working with facet_nodes()
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## **27.7 Network Metrics Analysis**

```{r}
#27.7.1 Computing centrality indices
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

```{r}
#27.7.2 Visualising network metrics
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

```{r}
#| warning: false
#27.7.3 Visualising Community
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight, 
      directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(
    aes(
      width=Weight), 
    alpha=0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))  

g + theme_graph()

g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y, 
        group = community, 
        fill = community),  
    alpha = 0.2,  
    expand = unit(0.3, "cm"),  # Expand
    radius = unit(0.3, "cm")  # Smoothness
  ) + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)
  
g + theme_graph()
```

## **27.8 Building Interactive Network Graph with visNetwork**

```{r}
#27.8.1 Data preparation
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

```{r}
#27.8.2 Plotting the first interactive network graph
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

```{r}
#27.8.3 Working with layout
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

# **29  Visualising and Analysing Text Data with R: tidytext methods**

```{r}
#29.2 Getting Started
pacman::p_load(tidytext, widyr, wordcloud, DT, ggwordcloud, textplot, lubridate, hms,
tidyverse, tidygraph, ggraph, igraph)
```

```{r}
#29.3 Importing Multiple Text Files from Multiple Folders
#29.3.1 Creating a folder list
news20 <- "data/20news/"
#29.3.2 Define a function to read all files from a folder into a data frame
read_folder <- function(infolder) {
  tibble(file = dir(infolder, 
                    full.names = TRUE)) %>%
    mutate(text = map(file, 
                      read_lines)) %>%
    transmute(id = basename(file), 
              text) %>%
    unnest(text)
}
```

## **29.4 Importing Multiple Text Files from Multiple Folders**

```{r}
#29.4.1 Reading in all the messages from the 20news folder
raw_text <- tibble(folder = dir(news20, full.names = TRUE)) %>%
  mutate(folder_out = map(folder, read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  transmute(newsgroup = basename(folder), id, text)

dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(raw_text, "data/rds/news20.rds")
```

```{r}
#| eval: false
#install.packages("ragg")
```

## **29.5 Initial EDA**

```{r}
raw_text <- read_rds("data/rds/news20.rds")
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```

## **29.6 Introducing tidytext**

```{r}
#29.6.1 Removing header and automated email signitures
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

```{r}
#29.6.2 Removing lines with nested text representing quotes from other users.
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text, 
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text, 
                     "^In article <")
  )
```

```{r}
#29.6.3 Text Data Processing
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)

usenet_words %>%
  count(word, sort = TRUE)

words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()
```

```{r, warning=FALSE}
#29.6.4 Visualising Words in newsgroups
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```

```{r, warning=FALSE}
#29.6.5 Visualising Words in newsgroups
set.seed(1234)

words_by_newsgroup %>%
  filter(n > 0) %>%
ggplot(aes(label = word,
           size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

## **29.7 Basic Concept of TF-IDF**

```{r}
#29.7.1 Computing tf-idf within newsgroups
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

```{r}
#29.7.3 Visualising tf-idf as interactive table
DT::datatable(tf_idf, filter = 'top') %>% 
  formatRound(columns = c('tf', 'idf', 
                          'tf_idf'), 
              digits = 3) %>%
  formatStyle(0, 
              target = 'row', 
              lineHeight='25%')
```

```{r}
#29.7.4 Visualising tf-idf within newsgroups
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, 
            n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word, 
                        tf_idf)) %>%
  ggplot(aes(tf_idf, 
             word, 
             fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup, 
             scales = "free") +
  labs(x = "tf-idf", 
       y = NULL)
```

```{r}
#29.7.5 Counting and correlating pairs of words with the widyr package
newsgroup_cors <- words_by_newsgroup %>%
  pairwise_cor(newsgroup, 
               word, 
               n, 
               sort = TRUE)
```

```{r}
#29.7.6 Visualising correlation as a network
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation, 
                     width = correlation)) +
  geom_node_point(size = 6, 
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 color = "red",
                 repel = TRUE) +
  theme_void()
```

```{r}
#29.7.7 Bigram
bigrams <- cleaned_text %>%
  unnest_tokens(bigram, 
                text, 
                token = "ngrams", 
                n = 2)
bigrams
```

### **29.7.8 Counting bigrams**

```{r}
bigrams_count <- bigrams %>%
  filter(bigram != 'NA') %>%
  count(bigram, sort = TRUE)
bigrams_count
```

### **29.7.9 Cleaning bigram**

```{r}
bigrams_separated <- bigrams %>%
  filter(bigram != 'NA') %>%
  separate(bigram, c("word1", "word2"), 
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)
bigrams_filtered
```

### **29.7.10 Counting the bigram again**

```{r}
bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE)
```

### **29.7.11 Create a network graph from bigram data frame**

```{r}
bigram_graph <- bigram_counts %>%
  filter(n > 3) %>%
  graph_from_data_frame()
bigram_graph
```

### **29.7.12 Visualizing a network of bigrams with ggraph**

```{r}
set.seed(1234)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name), 
                 vjust = 1, 
                 hjust = 1)
```

### **29.7.13 Revised version**

```{r}
set.seed(1234)

a <- grid::arrow(type = "closed", 
                 length = unit(.15,
                               "inches"))

ggraph(bigram_graph, 
       layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), 
                 show.legend = FALSE,
                 arrow = a, 
                 end_cap = circle(.07,
                                  'inches')) +
  geom_node_point(color = "lightblue", 
                  size = 5) +
  geom_node_text(aes(label = name), 
                 vjust = 1, 
                 hjust = 1) +
  theme_void()
```
